Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-01-07 versionForJCGS/PaperJCGS.Rnw
changes for submission

%-----------------------------------------------------------------------------------------------------
%%% Title
%-----------------------------------------------------------------------------------------------------
\if0\blind
{
%opening
\title{Binning Strategies and Related Loss for Large Data}
\author{Karsten Maurer$^1$, Heike Hofmann$^{2,3}$, Susan VanderPlas$^2$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
  \maketitle
} \fi

\if1\blind
{
\title{Binning Strategies and Related Loss for Large Data}
\author{Author A$^1$, Author B$^{2,3}$, Author C$^2$\\$^1$Affiliation X \\$^2$Affiliation Y \\$^3$Affiliation Z}
  \maketitle
} \fi

@@

\noindent%
{\it Keywords:}  scatterplots, visual loss, aggregation, graphics
\vfill

@@

p2 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor)) + 
  scale_colour_manual(values = c("darkblue","red"))+
  geom_point(aes(x=x,y=y), size=I(3))+ 
  theme_bw() + xlab("") + ylab("") +
  ggtitle("Random Binning") +
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"),
            aspect.ratio=1)

@@

p3 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor)) + 
  scale_colour_manual(values = c("darkblue","red"))+
  geom_point(aes(x=x,y=y), size=I(3))+ 
  theme_bw() + xlab("") + ylab("") +
  ggtitle("Post-Processed Random Binning") +
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"),
            aspect.ratio=1)

@@

The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}. We see that in random binning there are pairs of points -- like those with line segments highlighted in red -- that have been allocated in such a way that they are closer to their partner's bin center than their own. We see that the binned scatterplot remains identical if these points are swapped back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed. To appropriately reflect the perceived loss of spatial information we use the \textit{net spatial loss}, $L^S_{net}$, which is the minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot. For standard binned data, the net spatial loss is always equivalent to the total spatial loss due to the deterministic bin allocations. For random binned data, the net spatial loss is achieved by first exchanging bin assignments for all pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then calculating the total spatial loss from this post-processed binned data.

>>>>

%-----------------------------------------------------------------------------------------------------
%%% Title
%-----------------------------------------------------------------------------------------------------
\if0\blind
{
%opening
\title{Binning Strategies and Related Loss for Binned Scatterplots of Large Data}
\author{Karsten Maurer$^1$, Heike Hofmann$^{2,3}$, Susan VanderPlas$^2$\\$^1$Department of Statistics, Miami University \\$^2$Department of Statistics, Iowa State University \\$^3$Human Computer Interaction, Iowa State University}
  \maketitle
} \fi

\if1\blind
{
\title{Binning Strategies and Related Loss for Binned Scatterplots of Large Data}
\author{Author A$^1$, Author B$^{2,3}$, Author C$^2$\\$^1$Affiliation X \\$^2$Affiliation Y \\$^3$Affiliation Z}
  \maketitle
} \fi

@@

\noindent%
{\it Keywords:}  binned scatterplots, visual loss, aggregation, graphics
\vfill

@@

p2 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
  geom_segment(aes(x = x, y = y, xend = rxcent, yend = rycent, color=linecolor, linetype=linecolor)) + 
  scale_colour_manual(values = c("darkblue","red"))+
  geom_point(aes(x=x,y=y), size=I(3))+ 
  theme_bw() + xlab("") + ylab("") +
  ggtitle("Random Binning") +
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"),
            aspect.ratio=1)

@@

p3 <- qplot(xcent, ycent, geom="point", size=I(5), shape=I(3), color=I("gray")) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), data=temp) + 
  geom_segment(aes(x = x, y = y, xend = rxcentpost, yend = rycentpost, color=linecolor, linetype=linecolor)) + 
  scale_colour_manual(values = c("darkblue","red"))+
  geom_point(aes(x=x,y=y), size=I(3))+ 
  theme_bw() + xlab("") + ylab("") +
  ggtitle("Post-Processed Random Binning") +
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank(),
        legend.position="none",plot.margin = unit(c(0,1,0, 1),"cm"),
            aspect.ratio=1)

@@

The total spatial loss for randomly binned data is visualized in Figure~\ref{fig:spatlossdemo2}. We see that in random binning there are pairs of points -- like those with dashed line segments -- that have been allocated in such a way that they are closer to their partner's bin center than their own. We see that the binned scatterplot remains identical if these points are swapped back to the closer centers; however, the total spatial loss is smaller after the random allocation is post-processed. To appropriately reflect the perceived loss of spatial information we use the \textit{net spatial loss}, $L^S_{net}$, which is the minimum total spatial loss from all binning allocations that result in the same reduced binned data, and thus the same binned scatterplot. For standard binned data, the net spatial loss is always equivalent to the total spatial loss due to the deterministic bin allocations. For random binned data, the net spatial loss is achieved by first exchanging bin assignments for all pairs of points in neighboring bins that exist further from their own bin center than from their partner's bin center, then calculating the total spatial loss from this post-processed binned data.
